---
title: "Machine Learning Project"
date: "Tuesday, February 17, 2015"
output: html_document
---
The goal of this project is to predict the manner in which participants did the exercise. This is the "classe" variable in the training set.

Begin by reading in the provided training dataset.
```{r}
library(caret)
library(rattle)
library(rpart)
library(randomForest)
train <- read.csv("data/pml-training.csv", header=TRUE, na.strings="NA")
```
Pre-process by removing rows with new_window="yes" as well as empty and NA columns. 
```{r}
train <- train[train$new_window!="yes",]
train <- train[,c(8:11, 37:49, 60:68, 84:86, 102, 113:124, 140, 151:160)]
dim(train)
```
For cross-validation, split into training and testing
```{r}
inTrain <- createDataPartition(y=train$classe, p=0.75, list=FALSE)
training <- train[inTrain,]
testing <- train[-inTrain,]
dim(training)
dim(testing)
```
Since there are so many variables and it's difficult to pick one as the predictor, I use the random forest model on the training dataset.
```{r}
rf <- randomForest(formula = classe ~ ., data = training, ntree = 500) 
rf
```
Run a prediction on the cross-validation testing dataset and show predicted results.
```{r}
pred <- predict(rf, testing)
table(pred, testing$classe)
```
Calculate the predicted out of sample error rate
```{r}
testing$predRight <- pred==testing$classe
predictedOutOfSampleErrorRate <- table(testing$predRight)[1] / table(testing$predRight)[2] 
predictedOutOfSampleErrorRate
```
Now read the testing dataset that was provided and pre-process exactly as was done for the training dataset.
```{r}
testing <- read.csv("data/pml-testing.csv", header=TRUE, na.strings="NA")
testing <- testing[testing$new_window!="yes",]
testing <- testing[,c(8:11, 37:49, 60:68, 84:86, 102, 113:124, 140, 151:159)]
```
Add the classe factor column so that the structure of testing exactly matches training for a later prediction.
```{r}
testing$classe <- ""
testing$classe <- factor(testing$classe)
dim(testing)
```
Run prediction on the testing dataset
```{r}
pred <- predict(rf, testing)
table(pred, testing$classe)
```
Create the files to be submitted for submission
```{r}
submitPred <- as.character(pred)
n = length(submitPred)
for(i in 1:n){
  filename = paste0("problem_id_",i,".txt")
  write.table(submitPred[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
```