---
title: "Machine Learning Project"
date: "Tuesday, February 18, 2015"
output: html_document
---
The goal of this project is to predict the manner in which participants did the exercise. This is the "classe" variable in the training set.
```{r include=FALSE}
library(caret)
library(rattle)
library(rpart)
library(randomForest)
```

Begin by loading the libraries and reading the provided training dataset.
```{r}
train <- read.csv("data/pml-training.csv", header=TRUE, na.strings="NA")
```
### Pre-Processing

Pre-process by removing rows with new_window="yes" as well as empty and NA columns. 
```{r}
train <- train[train$new_window!="yes",]
train <- train[,c(8:11, 37:49, 60:68, 84:86, 102, 113:124, 140, 151:160)]
dim(train)
```
### Cross-Validation

For cross-validation, split into training and testing
```{r}
inTrain <- createDataPartition(y=train$classe, p=0.75, list=FALSE)
training <- train[inTrain,]
testing <- train[-inTrain,]
dim(training)
dim(testing)
```
### Use Random Forest Model

Since there are so many variables and it's difficult to pick one as the predictor, I use the random forest model on the training dataset.
```{r}
rf <- randomForest(formula = classe ~ ., data = training, ntree = 500) 
rf
```
### Run Prediction on cross-validation dataset

Run a prediction on the cross-validation testing dataset and show predicted results.
```{r}
pred <- predict(rf, testing)
table(pred, testing$classe)
```

### Calculate Out of Sample Error Rate

Calculate the predicted out of sample error rate.
```{r}
testing$predRight <- pred==testing$classe
predictedOutOfSampleErrorRate <- table(testing$predRight)[1] / length(testing$predRight)
predictedOutOfSampleErrorRate
```
The predicted out of sample error rate is `r predictedOutOfSampleErrorRate`.

### Run Prediction on Testing Dataset

Now read the testing dataset that was provided and pre-process exactly as was done for the training dataset.
```{r}
testing <- read.csv("data/pml-testing.csv", header=TRUE, na.strings="NA")
testing <- testing[testing$new_window!="yes",]
testing <- testing[,c(8:11, 37:49, 60:68, 84:86, 102, 113:124, 140, 151:159)]
```
Add the classe factor column so that the structure of testing exactly matches training for a later prediction.
```{r}
testing$classe <- ""
testing$classe <- factor(testing$classe)
dim(testing)
```
Run prediction on the testing dataset and show results.
```{r}
pred <- predict(rf, testing)
table(pred, testing$classe)
```
### Create Files for Grading

Create the files to be submitted for grading.
```{r}
submitPred <- as.character(pred)
n = length(submitPred)
for(i in 1:n){
  filename = paste0("problem_id_",i,".txt")
  write.table(submitPred[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
```